---
title: "Project 1, Part 2"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due dates:** Friday, February 17 at 12pm (noon) on Canvas. Submit *both* your knitted HTML file and the R Markdown file to Canvas.

**Collaboration:** This project *must* be completed independently. See "Collaboration" below for more details.

# Overview

In the first part of this project, you began to explore the research questions with exploratory data analysis for the relevant variables. In the second part of this project, you will fit a model to address the **first** research question, perform model diagnostics, and conduct a hypothesis test.

# Data

The Behavioral Risk Factor Surveillance System (BRFSS) is an annual telephone survey of 350,000 people in the United States. As its name implies, the BRFSS is designed to identify risk factors in the adult population and report emerging health trends. For example, respondents are asked about their diet and weekly physical activity, their HIV/AIDS status, possible tobacco use, and even their level of healthcare coverage. The [BRFSS Web site](https://www.cdc.gov/brfss/) contains a complete description of the survey, including the research questions that motivate the study and many interesting results derived from the data.

We will focus on a random sample of 20,000 people from the BRFSS survey. While there are over 200 variables in this data set, we will work with a small subset for this project.

## Variables

The dataset provided for this project contains the following columns:

* `genhlth`: respondents were asked to evaluate their general health, responding either excellent, very good, good, fair or poor.
* `exerany`: indicates whether the respondent exercised in the past month (1) or did not (0).
* `hlthplan`: indicates whether the respondent had some form of health coverage (1) or did not (0).
* `smoke100`: indicates whether the respondent had smoked at least 100 cigarettes in their lifetime.
* `height`: in inches
* `weight`: in pounds
* `wtdesire`: desired weight in pounds
* `age`: in years
* `gender`: biological sex, limited to male/female.

## Downloading the data

To load the data, use the code below. It will import a data set called `cdc` into R.

```r
source("http://www.openintro.org/stat/data/cdc.R")
```

This code is included in the setup chunk of the R Markdown template for the project.

# Research questions

You are asked by a team of researchers to investigate respondents' exercise habits. For this part of the project, you will focus **only** on the first research question:

1. Is there a relationship between how much weight someone wants to lose, and the probability that they exercise regularly, after accounting for their age, general health, and health coverage?


# Part 2 overview

For the second part of the project, you will fit a model, perform model diagnostics, and use the model to address the first research question. You will submit two documents on Canvas for part 2:

* Your written report (.html file):
  * This is the knitted write up that will explain your work, and the answers to the researchers' questions
  * In your formal report, there should be **no code** showing. This
includes warnings and other stray code output – hide it all.
  * You will be graded on writing as well as your statistical analysis.
* Your R Markdown (.Rmd) file, with all the code you used for your analysis
  * The goal is that a person who reads your report, and wants to
replicate your results, could look at your .Rmd file and completely reproduce the results and figures in the knitted report.

# Part 2 content

Your written report for part 2 will contain three sections: EDA with empirical logit plots, model fitting and diagnostics, and hypothesis testing. Each section should be written in paragraph form, and make sure to always discuss any figures in your report.

### Exploratory data analysis: empirical logit plots

Use empirical logit plots to examine the relationship between quantitative explanatory variables and your response variable of interest. 

* You should show empirical logit plots for at least **2** quantitative explanatory variables. 
* Discuss each plot, and whether the shape assumption appears reasonable for that variable or if a transformation is needed. 
* If a transformation is needed, show an empirical logit plot for the transformed variable which demonstrates an improvement in the shape assumption

### Modeling and diagnostics

Using the explanatory variables requested for research question 1, build a model to address the first research question.

* Report the population model and the equation of the fitted model, with appropriate notation
* Use any transformations you deemed necessary from your empirical logit plots

Then, perform model diagnostics with your fitted model:

* Create quantile residual plots to assess the shape assumption
* Check for collinearity and influential points
* If any assumptions are still violated, modify your model and report the final model

Finally, interpret any relevant fitted coefficients in context of the research question.

### Hypothesis testing

Perform a hypothesis test that allows you to address the first question. When testing hypotheses, make sure to:

* State the hypotheses in terms of one or more model parameters
* Calculate a test statistic and p-value
* Use the p-value to make a conclusion about the research question

# Appearance and style

### Writing

The report should be written like an article or research paper: in full sentences and paragraphs, with headings for each section. You should not write your report with question numbers or as a list of bullet points. Scientific articles are generally written in third person, though "we" can also be acceptable ("we can see from Figure 2.1 ...") in some disciplines.

### Code

As with previous assignments, you will write the document in R Markdown. However, in full reports, the only output that should be visible from code chunks are figures and tables. If a code chunk does not produce a figure or table, you can hide it from the knitted document with `include=F`:

````
```{r, include=F}`r ''`

```
````

If a code chunk produces a figure or table, only the figure or table should be visible in the knitted document. You can hide the chunk but display the output with `echo=F, message=F, warning=F`:

````
```{r, echo=F, message=F, warning=F}`r ''`

```
````

### Figures

Figures should have labeled axes, and should be clear and easy to read. Figures should also be captioned and numbered; to caption a figure, use `fig.cap = "..."` in the chunk options. For example (scroll to the right on the code to see it all),

````
```{r, echo=F, message=F, warning=F, fig.cap="Figure 2.1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}`r ''`
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```
````

is displayed as

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```

Captions should provide enough information to understand what is being plotted, but interpretation can be left to the main text. Refer to figures by their number in the text. Make sure that any figures you include are discussed in the text.

### Tables

Tables should be nicely formatted, and have a number and caption. This can be done with the `kable` function in the `knitr` package.

For example,

````
```{r, echo=F, message=F}`r ''`
table(penguins$island, penguins$species) %>%
  knitr::kable(caption = "Table 3.2: Penguins by island and species")
```
````

is displayed as

```{r, echo=F, message=F}
table(penguins$island, penguins$species) %>%
  knitr::kable(caption = "Table 3.2: Penguins by island and species")
```

### Writing math in R Markdown

If you want to write mathematical notation, we need to tell Markdown, “Hey, we’re going to make a math symbol!” To do that, you use dollar signs. For instance, to make $\widehat{\beta}_1$, you simply put `$\widehathat{\beta}_1$` into the white space (not a chunk) in your Markdown.

Here are some examples of writing math, which you can adapt:

| Math | Code |
| --- | --- |
| $Y_i \sim Bernoulli(\pi_i)$ | `$Y_i \sim Bernoulli(\pi_i)$` |
| $\log \left( \dfrac{\pi_i}{1 - \pi_i} \right)$ | `$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right)$` |
| $\widehat{\pi}_i$ | `$\widehat{\pi}_i$` |

# Collaboration

One purpose of the projects is to capture your knowledge and understanding of statistics and data analysis. Unlike homework and lab assignments, this project must be completed on your own, without collaborating with other students. 

You may:

* ask the instructor for help
* use resources from the class notes and internet (Wikipedia, StackExchange, etc.)

You may not:

* work with another student on the project
* ask a TA or tutor (e.g., at the MSC) for help