---
title: "HW 3 Solutions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

titanic <- read.csv("https://sta214-s23.github.io/homework/Titanic.csv")
titanic <- titanic %>%
  drop_na()
```

## Question 1 (2 pts)

The null hypothesis is $H_0: \beta_1 = 0$, and the alternative hypothesis is $H_A: \beta_1 \neq 0$.

**Grading:** 1 pt for null, 1 pt for alternative. Lose 1 pt if not written in terms of $\beta_1$.

## Question 2 (4 pts)

Full model:

$Survived_i \sim Bernoulli(\pi_i)$

$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 Age_i$

Reduced model:

$Survived_i \sim Bernoulli(\pi_i)$

$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0$

**Grading:** 2 pts for full model, 2 pts for reduced model. Lose 1 pt for incorrect notation (missing subscripts, adding hats, etc.) Lose 1 pt if only the systematic component is included (missing random component).

## Question 3

### (a) (3 pts)

```{r}
m2 <- glm(Survived ~ 1, data = titanic, family = binomial)
summary(m2)
```

$Survived_i \sim Bernoulli(\pi_i)$

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -0.38$

**Grading:** Lose 1 pt if missing code, lose 2 pts if missing equation of the fitted reduced model. Lose 1 pt for incorrect notation (missing subscripts, hats, etc.)

### (b) (1 pt)

The deviance of the reduced model is 964.52

**Grading:** it is ok if they round slightly. If they didn't remove missing values before fitting the model, they would get 1186.7 instead. If they don't remove missing values, lose 1 pt.

### (c) (3 pts)

```{r}
m1 <- glm(Survived ~ Age, data = titanic, family = binomial)
summary(m1)
```

$Survived_i \sim Bernoulli(\pi_i)$

$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -0.057 - 0.011 Age_i$

**Grading:** Lose 1 pt if missing code, lose 2 pts if missing equation of the fitted full model. Lose 1 pt for incorrect notation (missing subscripts, hats, etc.)

### (d) (1 pt)

The deviance of the full model is 960.23


## Question 4 (2 pts)

In this case, the reduced model *is* the intercept-only model. Therefore, the deviance of the reduced model agrees with the null deviance.

## Question 5 (1 pt)

$G = 964.52 - 960.23 = 4.29$

**Grading:** it is ok if they round slightly

## Question 6

### (a) (3 pts)

```{r}
nsim <- 500
null_statistics <- rep(0, nsim)
for(i in 1:nsim){
  x <- titanic$Age 
  p <- exp(-0.38 + 0*x)/(1 + exp(-0.38 + 0*x))
  y <- rbinom(length(x), 1, p)
  m1 <- glm(y ~ x, family = binomial)
  
  null_statistics[i] <- m1$null.deviance - 
    m1$deviance
}
hist(null_statistics)
```

**Grading:** Lose 1 pt for not changing the explanatory variable `x` (should be Age), lose 1 pt for not changing the intercept in the probabilities (should be -0.38).

### (b) (2 pts)

The observed test statistic was $G = 4.29$, which looks reasonably unusual compared to our null statistics. Most simulated statistics from the null distribution are below 4 or so.

**Grading:** 1 pt for correct answer, 1 pt for explaining reasoning

### (c) (2 pts)

```{r}
mean(null_statistics >= 4.29)
```

**Grading:** 1 pt for showing code, 1 pt for correctly using observed drop in deviance

## Question 7 (2 pts)

```{r}
x <- seq(0, 1, 0.1)
x_cubed <- rep(0, length(x))
for(i in 1:length(x)){
  x_cubed[i] <- x[i]^3 
}
x_cubed
```

**Grading:** 2 pts for using `x[i]^3`. They don't need to rename the vector `x_cubed`, but it would be helpful.


## Question 8 (2 pts)

```{r}
x <- seq(0, 2, 0.05)
x_cubed <- rep(0, length(x))
for(i in 1:length(x)){
  x_cubed[i] <- x[i]^3 
}
x_cubed
```

**Grading:** They need to change both the range of `x` (0 to 2), and the increment (by 0.05). It is fine for them to calculate either $x^2$ or $x^3$


## Question 9

### (a) (2 pts)

```{r}
n <- 1000
normal_sample <- rep(0, n)
for(i in 1:n){
  normal_sample[i] <- rnorm(1, 0, 1)
}
hist(normal_sample)
```

### (b) (1 pt)

Yes, the histogram looks like a normal distribution (unimodal, symmetric, bell-shaped, etc.)

**Grading:** 1 pt for correct answer.

## Question 10

### (a) (2 pts)

```{r}
n <- 2000
chisq_sample <- rep(0, n)
for(i in 1:n){
  chisq_sample[i] <- rchisq(1, df=1)
}
hist(chisq_sample)
```

**Grading:** They need to change both `n` and the distribution they are sampling from. They don't need to change the name of the vector, but it would be helpful.

### (b) (1 pt)

It looks pretty similar!

**Grading:** 1 pt for correct answer

## Question 11 (Extra credit)

They can earn 1 pt of extra credit for each part of this question which they answer correctly.

### (a) (1 pt)

The issue is that `pi0` has length 11, but `likelihood` has length 12. When we go to access `pi0[12]`, there is nothing there!

### (b) (1 pt)

```{r}
pi0 <- seq(0, 1, 0.1)
likelihood <- rep(0, 12)
for(i in 1:11){
  likelihood[i] <- pi0[i]^2 * (1 - pi0[i])^3
}
likelihood
```