---
title: Hypothesis testing for logistic regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, include=F}
library(tidyverse)
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

### Modeling grad admissions

.large[
**Data:** Grad application data
  * `admit`: accepted to grad school? (0 = no, 1 = yes)
  * `gre`: GRE score
  * `gpa`: undergrad GPA
  * `rank`: prestige of undergrad institution
  
**New question:** Does the relationship between GPA and the probability of acceptance depend on the prestige of a student's undergrad institution, after accounting for GRE score?
  
.question[
How could we use hypothesis testing to investigate this research question? Discuss with your neighbor, then we will discuss as a group.
]
]

---

### Model

.large[
$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$$
$$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$$

.question[
How would I interpret coefficients when I have an interaction?
]
]

---

### Hypotheses

.large[
$$Admit_i \sim Bernoulli(\pi_i)$$

$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$$
$$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$$

**Question:** Does the relationship between GPA and the probability of acceptance depend on the prestige of a student's undergrad institution, after accounting for GRE score?

.question[
What are my null and alternative hypotheses?
]
]

---

### Class activity, Part I

.large[
[https://sta214-s23.github.io/class_activities/ca_lecture_9.html](https://sta214-s23.github.io/class_activities/ca_lecture_9.html)
]

---

### Class activity

.large[
**Models:**

```{r}
m1 <- glm(admit ~ gre + gpa*as.factor(rank),
          data = grad_app, family = binomial)

m2 <- glm(admit ~ gre + gpa + as.factor(rank),
          data = grad_app, family = binomial)

m2$deviance - m1$deviance
```
]

---

### Class activity

.large[
$H_0: \beta_6 = \beta_7 = \beta_8 = 0$

**Test statistic:** $G = 0.41$

**Calculating a p-value:**

```{r, eval=F}
pchisq(..., ..., lower.tail=F)
```
]

---

### Class activity

.large[
$H_0: \beta_6 = \beta_7 = \beta_8 = 0$

**Test statistic:** $G = 0.41$

**Calculating a p-value:**

```{r}
pchisq(0.41, 3, lower.tail=F)
```
]

---

### Likelihood ratio test for nested models

.large[
**Goal:** Compare full and reduced models

**Steps:**

* Calculate deviance for full and reduced models
* $G$ = deviance for reduced - deviance for full
* p-value: $G \sim \chi^2_{q}$
]

---

### Alternative: Wald tests for single parameters

.large[
$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$
$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$

**Hypotheses:**

<br/>

**Test statistic:**

$z =$
]

---

### Example

.large[
$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$
$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$

$H_0: \beta_1 = 0 \hspace{1cm} H_A: \beta_1 \neq 0$
]

```{r, echo=F, output.lines=10:14, highlight.output=c(5)}
grad_glm <- glm(admit ~ gre + gpa*as.factor(rank), data = grad_app, family = binomial)
summary(grad_glm)
```

.large[
$z =$
]

---

### Example

.large[
$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$
$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$

$H_0: \beta_1 = 0 \hspace{1cm} H_A: \beta_1 > 0$
]

```{r, echo=F, output.lines=10:14, highlight.output=c(5)}
grad_glm <- glm(admit ~ gre + gpa*as.factor(rank), data = grad_app, family = binomial)
summary(grad_glm)
```

.large[
$z =$

.question[
How do I calculate a p-value?
]
]

---

### Example

.large[
$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$
$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$

$H_0: \beta_1 = 0 \hspace{1cm} H_A: \beta_1 > 0$
]

```{r, echo=F, output.lines=10:14, highlight.output=c(5)}
grad_glm <- glm(admit ~ gre + gpa*as.factor(rank), data = grad_app, family = binomial)
summary(grad_glm)
```

.large[

```{r}
pnorm(2.075, lower.tail=F)
```
]

---

### Wald tests vs. likelihood ratio tests

.large[
.pull-left[
**Wald test**

* like t-tests
* test a single parameter
* some example hypotheses:
    * $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$
    * $H_0: \beta_1 = 1$ vs. $H_A: \beta_1 > 1$
]

.pull-right[
**Likelihood ratio test**

* like nested F-tests
* test one or more parameters 
* some example hypotheses:
    * $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$
]

p-values are different, because test statistics and distributions are different
]

---

### Confidence intervals

.large[
$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$
$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$

Now we want a "reasonable range" of values for $\beta_1$.

**Confidence interval:**
]

---

### Computing $z^*$

.large[
Example: for a 95% confidence interval, $z^* = 1.96$

```{r}
qnorm(0.025, lower.tail=F)
```

Example: for a 99% confidence interval, $z^* = 2.58$:

```{r}
qnorm(0.005, lower.tail=F)
```
]

---

### Confidence interval

```{r, echo=F, output.lines=10:14, highlight.output=c(5)}
grad_glm <- glm(admit ~ gre + gpa*as.factor(rank), data = grad_app, family = binomial)
summary(grad_glm)
```

.large[
95% confidence interval for $\beta_1$:
]

---

### Class activity, Part II

.large[
[https://sta214-s23.github.io/class_activities/ca_lecture_9.html](https://sta214-s23.github.io/class_activities/ca_lecture_9.html)
]

---

### Class activity

.large[
```{r, include=F}
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

m1 <- glm(Dengue ~ WBC, data = dengue, family = binomial)
```

```{r, echo=F, output.lines=10:14}
summary(m1)
```
]