---
title: Logistic regression review
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
earthquake <- read.csv("https://sta214-s23.github.io/class_activities/earthquake_small.csv")

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

### Class activity: Question 2

```{r, output.lines = 10:17}
m1 <- glm(Damage ~ age*land_surface_condition, 
          family = binomial, data = earthquake)
summary(m1)
```

---

### Class activity: Question 3

```{r, echo=F, output.lines = 10:17}
summary(m1)
```

---

### Class activity: Question 4

```{r, echo=F, output.lines = 10:17}
summary(m1)
```

---

### Class activity: Question 5

.large[
$H_0:$

$H_A:$
]

---

### Class activity: Question 6

```{r}
m1 <- glm(Damage ~ age*land_surface_condition, 
          family = binomial, data = earthquake)
m2 <- glm(Damage ~ age + land_surface_condition, 
          family = binomial, data = earthquake)

m2$deviance - m1$deviance

pchisq(5.12, df=2, lower.tail=F)
```

---

### Class activity: Question 7

```{r, echo=F, output.lines = 10:17}
summary(m1)
```

---

### Class activity: Question 8

.large[
```{r, echo=F, message=F, fig.align='center', fig.width=8, fig.height=6}
library(statmod)
earthquake %>%
  mutate(resids = qresid(m1)) %>%
  ggplot(aes(x = age, y = resids)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Age", y = "Quantile residuals") +
  theme_bw() +
  theme(text = element_text(size = 20))
```
]

---

### Class activity: Question 10

.large[
```{r}
table(Prediction = m1$fitted.values > 0.85, 
      Truth = earthquake$Damage)
```
]

---

### Class activity: Question 11

```{r, message=F, warning=F, output.lines = 10:25}
library(MASS)
m0 <- glm(Damage ~ 1, family = binomial, data = earthquake)
forward_aic <- stepAIC(m0, scope = ~ age + area_percentage + 
                         height_percentage + land_surface_condition + 
                         foundation_type + roof_type + 
                         count_families + has_secondary_use,
                       trace = 0, direction = "forward")
summary(forward_aic)
```

---

### Class activity: Question 12

.large[
.question[
Why can't we use a hypothesis test to compare the results of forward selection with our original model?
]
]

---

### Class activity: Question 13

.large[
```{r, echo=F, message=F, fig.align='center', fig.width=10, fig.height=4}
library(ROCR)
library(gridExtra)

pred <- prediction(m1$fitted.values, earthquake$Damage)
perf <- performance(pred,"tpr","fpr")
#performance(pred, "auc")@y.values
p1 <- data.frame(fpr = perf@x.values[[1]], 
           tpr = perf@y.values[[1]]) %>%
  ggplot(aes(x = fpr, y = tpr)) +
  geom_line(lwd=1.5) +
  geom_abline(slope = 1, intercept = 0, lty = 2,
  lwd = 1.5) +
  labs(x = "False positive rate (1 - Specificity)", 
       y = "True positive rate (Sensitivity)", 
       title = "Original model (AUC = 0.73)") +
  theme_classic()

pred <- prediction(forward_aic$fitted.values, earthquake$Damage)
perf <- performance(pred,"tpr","fpr")
#performance(pred, "auc")@y.values
p2 <- data.frame(fpr = perf@x.values[[1]], 
           tpr = perf@y.values[[1]]) %>%
  ggplot(aes(x = fpr, y = tpr)) +
  geom_line(lwd=1.5) +
  geom_abline(slope = 1, intercept = 0, lty = 2,
  lwd = 1.5) +
  labs(x = "False positive rate (1 - Specificity)", 
       y = "True positive rate (Sensitivity)", 
       title = "Model from stepwise selection (AUC = 0.82)") +
  theme_classic()

grid.arrange(p1, p2, ncol=2)
```
]




