---
title: Poisson Regression Inference
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(knitr)
library(tidyverse)

crimes <- read_csv("~/Documents/Teaching/sta214-f22.github.io/slides/c_data.csv")

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

m1 <- glm(nv ~ region, data = crimes, family = poisson)
```

### STA courses next semester

.large[
Some classes to consider after STA 214:

* STA 247 Design and Sampling
* STA 279 Statistical Computing
* STA 310 Probability (requires calc II)
* STA 363 Intro to Statistical Learning (requires linear algebra)

Other cool courses to consider:

* STA 311 Statistical Inference (requires STA 310)
* STA 312 Linear Models (requires STA 310 and linear algebra)
* STA 362 Multivariate Statistics (requires linear algebra)
* STA 365 Applied Bayesian Statistics (requires STA 310)
* STA 368 Time Series and Forecasting (requires STA 310)
]

---

### Last time

.large[
A concerned parent asks us to investigate crime rates on college campuses. We have access to data on 81 different colleges and universities in the US, including the following variables:

* `type`: college (C) or university (U)
* `nv`: the number of crimes for that institution in the given year
* `enroll1000`: the number of enrolled students, in thousands
* `region`: region of the US C = Central, MW = Midwest, NE = Northeast, SE = Southeast, SW = Southwest, and W = West)

$$Crimes_i \sim Poisson(\lambda_i)$$

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i$$
]

---

### Goodness of fit

.large[
**Goodness of fit test:** If the model is a good fit for the data, then the residual deviance follows a $\chi^2$ distribution with the same degrees of freedom as the residual deviance

Residual deviance = 621.24, df = 75

```{r}
pchisq(621.24, df=75, lower.tail=F)
```

So our model might not be a very good fit to the data.

.question[
Why might our model not be a good fit?
]
]

---

### Potential issues with our model

.large[
* The Poisson distribution might not be a good choice
* There may be additional factors related to the number of crimes which we are not including in the model

.question[
Which other factors might be related to the number of crimes?
]
]

---

### Offsets

.large[
We will account for school size by including an **offset** in the model:

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$
]

---

### Motivation for offsets

.large[
We can rewrite our regression model with the offset:

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$
]

---

### Fitting a model with an offset

.large[
```{r, output.lines = 10:16}
m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)
```

* The offset doesn't show up in the output (because we're not estimating a coefficient for it)
]

---

### Fitting a model with an offset

.large[
$$\log(\widehat{\lambda}_i) = -1.30 + 0.10 MW_i + 0.76 NE_i + \\ 0.87 SE_i + 0.51 SW_i + 0.21 W_i \\  + \log(Enrollment_i)$$

.question[
How would I interpret the intercept -1.30?
]
]

---

### When to use offsets

.large[
Offsets are useful in Poisson regression when our counts come from groups of very different sizes (e.g., different numbers of students on a college campus). The offset lets us interpret model coefficients in terms of rates instead of raw counts.

.question[
With your neighbor, brainstorm some other data scenarios where our response is a count variable, and an offset would be useful. What would our offset be?
]
]

---

### Inference

.large[
$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$

Our concerned parent wants to know whether the crime rate on campuses is different in different regions. 

.question[
What hypotheses would we test to answer this question?
]
]

---

### Likelihood ratio test

.large[
Full model:

$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$

Reduced model:

$$\log(\lambda_i) = \beta_0 + \log(Enrollment_i)$$
]

---

### Likelihood ratio test

.large[
```{r, output.lines = 22:23}
m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)
```

.question[
What is my test statistic?
]
]

---

### Likelihood ratio test

.large[
```{r, output.lines = 22:23}
m2 <- glm(nv ~ region, offset = log(enroll1000),
          data = crimes, family = poisson)
summary(m2)
```

$G = 491 - 433.14 = 57.86$

```{r}
pchisq(57.86, df=5, lower.tail=F)
```
]

---

### Inference

.large[
$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$

Now our concerned parent wants to know about the difference between Western and Central schools. They would like a "reasonable range" of values for the difference between the regions.

.question[
How would we construct a "reasonable range" of values for this difference?
]
]

---

### Confidence intervals

.large[
$$\log(\lambda_i) = \beta_0 + \beta_1 MW_i + \beta_2 NE_i + \beta_3 SE_i + \beta_4 SW_i + \beta_5 W_i \\ + \log(Enrollment_i)$$

```{r, echo=F, output.lines = 10:16}
summary(m2)
```

95% confidence interval for $\beta_5$:
]

---

### Class activity

.large[
[https://sta214-s23.github.io/class_activities/ca_lecture_20.html](https://sta214-s23.github.io/class_activities/ca_lecture_20.html)
]

---

### Class activity

.large[

$$Articles_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Female_i + \beta_2 Married_i + \beta_3 Kids_i + \\  \beta_4 Prestige_i + \beta_5 Mentor_i$$

.question[
Do I need an offset for this model?
]
]

---

### Class activity

.large[
$$Articles_i \sim Poisson(\lambda_i)$$
$$\log(\lambda_i) = \beta_0 + \beta_1 Female_i + \beta_2 Married_i + \beta_3 Kids_i + \\  \beta_4 Prestige_i + \beta_5 Mentor_i$$

.question[
We are interested in the relationship between prestige and the number of articles published, after accounting for other factors. What confidence interval should we make?
]
]

---

### Class activity

.large[
```{r, echo=F, message=F, warning=F, output.lines = 10:16}
library(foreign)

articles <- read.dta("http://www.stata-press.com/data/lf2/couart2.dta")

m1 <- glm(art ~ ., data = articles, family = poisson)
summary(m1)
```

.question[
How do I construct a confidence interval for $\exp\{\beta_4\}$?
]
]

