---
title: "Class Activity, January 25"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

In this class activity, we will work with the grad application data from HW 1:

```{r, eval=F}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
```

On HW 1, you fit a logistic regression model on the grad application data with GPA as a predictor. Now let's consider GRE as a predictor instead. Our logistic regression model is

$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 \text{GRE}_i$

We are interested in testing $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$. Our null hypothesis corresponds to the reduced model 

$Y_i \sim Bernoulli(\pi_i) \hspace{0.5cm} \log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0$

The equation of the fitted reduced model is 

$$\log \left( \dfrac{\widehat{\pi}_i}{1 - \widehat{\pi}_i} \right) = -0.77$$

# Questions

To test these hypotheses, we want to do the following:

* Pretend the reduced model is correct
* Use the reduced model to simulate new data
* Calculate a test statistic for the simulated data
* Repeat the process many times to approximate the null distribution of the test statistic

On the slides, we wrote code to create a single simulated dataset from the reduced model, and calculate a test statistic:

```{r, eval=F}
x <- grad_app$gre
p <- exp(-0.77 + 0*x)/(1 + exp(-0.77 + 0*x))
y <- rbinom(length(x), 1, p)

m <- glm(y ~ x, family = binomial)
m$null.deviance - m$deviance
```

Now we want to repeat this process many times, and store the test statistics from each simulation.

1. Fill in the following `for` loop to create 500 simulated datasets and store the drop-in-deviance statistic $G$ for each simulation:

```{r, eval=F}
null_statistics <- c()
nsim <- ...
for(i in 1:nsim){
  ...
  
  null_statistics[i] <- ...
}
```

2. We can use a histogram to visualize the distribution of test statistics in the simulated data (i.e., the simulated null distribution):

```{r, eval=F}
hist(null_statistics)
```

3. Compared to the test statistics simulated from the reduced model, how unusual is the observed test statistic of $G = 13.92$?

