---
title: "Class Activity, February 1"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

# Part I

In the first part of this class activity, we will continue working with the grad application data from HW 1:

```{r, eval=F}
grad_app <- read.csv("https://stats.idre.ucla.edu/stat/data/binary.csv")
```

Our logistic regression model is

$$Admit_i \sim Bernoulli(\pi_i)$$
$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 GRE_i + \beta_2 GPA_i + \beta_3 Rank2_i \\ \hspace{1cm} + \beta_4 Rank3_i + \beta_5 Rank4_i + \beta_6 GPA_i \cdot Rank2_i$$
$$\hspace{3cm} + \beta_7 GPA_i \cdot Rank3_i + \beta_8 GPA_i \cdot Rank4_i$$

Our hypotheses are

$$H_0: \beta_6 = \beta_7 = \beta_8 = 0$$
$$H_A: \text{at least one of } \beta_6, \beta_7, \beta_8 \neq 0$$

1. Fit the full model in R, and calculate the deviance of the full model. *Remember to make `rank` a categorical variable in your model!*

2. Fit the reduced model in R, and calculate the deviance of the reduced model.

3. Calculate the drop-in-deviance test statistic $G$ comparing the reduced and full models.

4. Now we want to calculate a p-value. Rather than use parametric bootstrapping, we will use a $\chi^2$ distribution for the null distribution. Fill in the following code to calculate the p-value:

```{r, eval=F}
pchisq(..., ... , lower.tail=F)
```

# Part II

In the second part of this class activity, we will return to the dengue data from class. Our model is

$$Dengue_i \sim Bernoulli(\pi_i)$$
$$\log \left( \dfrac{\pi_i}{1 - \pi_i} \right) = \beta_0 + \beta_1 WBC_i$$

Below is the output from R:

```{r, echo=F}
dengue <- read.csv("https://sta279-s22.github.io/labs/dengue.csv")

m1 <- glm(Dengue ~ WBC, data = dengue, family = binomial)
summary(m1)
```

5. Calculate a 99% confidence interval for the change in the log odds of dengue associated with a one-unit increase in WBC.

6. How could we calculate a 99% confidence interval for the change in the *odds* of dengue associated with a one-unit increase in WBC?